---
title: "Practical Homework 1"
author: "Badamgarav Battushig"
date: "2026-04-19"
output: html_document
---

### Youth Substance Use Prediction using Tree-Based Models

#### Introduction 

This analysis uses the youth_data dataset, which contains youth demographic characteristics, experiences, and substance use information. The original dataset includes 10,561 observations and 79 variables. For modeling, a cleaned dataset was created using selected predictors and three target variables: Marijuana_use for binary classification, Marijuana_use_level for multi-class classification, and Alcohol_use_days for regression. After replacing non-substantive response codes with missing values and removing incomplete cases, the final modeling dataset contains 9,199 observations and 14 variables. For the binary marijuana-use outcome, the classes are imbalanced: 7,730 non-users (84.0%) and 1,469 users (16.0%). This imbalance is important because a model can appear accurate even if it mainly predicts the majority class.

#### Theoretical Background

Decision trees are supervised learning methods that divide the predictor space into smaller regions through recursive binary splitting. In classification, each split is chosen to make the resulting groups more homogeneous in terms of class membership, while in regression the splits are chosen to reduce variation within the response. A major strength of decision trees is interpretability, because the final model can be read as a sequence of simple decision rules. However, a single tree can be unstable and may overfit the training data, which is why pruning and tuning are important. The complexity parameter helps control tree size by removing unnecessary splits. Random forests extend this idea by averaging many trees built from bootstrap samples and random subsets of predictors, which reduces variance and often improves predictive performance. Boosting also combines many trees, but it builds them sequentially so that each new tree focuses on correcting previous errors. Because of these differences, tree-based methods provide a flexible set of tools for binary classification, multi-class classification, and regression.

#### Methodology

The data preparation process began by renaming the original dataframe and examining its size and structure. Special survey response codes such as 91, 93, 94, 97, and 98 were treated as non-substantive responses and replaced with NA so the models would not interpret them as meaningful numeric values. After selecting the response variables and predictors used in this homework, incomplete observations were removed with na.omit(), leaving 9,199 complete cases for analysis. For the binary classification task, the data were split into 70% training and 30% testing using createDataPartition(), resulting in 6,440 training cases and 2,759 testing cases. Model tuning was performed using 10-fold cross-validation. The decision tree and multi-class tree were tuned over a grid of cp values, the random forest was tuned over different mtry values, and the boosting model was tuned over combinations of number of trees, interaction depth, and shrinkage. This approach was used to choose model settings based on validation performance rather than arbitrary defaults.

### Data Type Comparison

The same set of demographic, behavioral, and social variables was used across all models. However, for each task, the response variable was excluded from the predictor set, resulting in slightly different input features depending on the modeling objective..

**Original dataset size: 10,561 rows and 79 columns**
**Final modeling dataset size: 9,199 rows and 14 columns**

---

### Load Required Libraries

```{r}
library(tidyverse)
library(caret)
library(rpart)
library(rpart.plot)
library(randomForest)
library(gbm)
library(dplyr)
```

---

### Load Dataset

```{r}
load("C:/Users/badam/OneDrive/Desktop/Data 5322 ML2/Practice HW 1/youth_data.Rdata")
head(df)
```

---

### Understanding the Dataset

The dataset `youth_data` is used as the main dataset for this analysis. It includes demographic variables, youth experiences, and substance use information.

```{r}
# Rename df to youth_data for clarity
youth_data <- df

# check dataset size
dim(youth_data)

# inspect structure
str(youth_data)

# summary statistics
summary(youth_data)

# column names
colnames(youth_data)
```

```{r}
# See variables in each group
demographic_cols
substance_cols
youth_experience_cols
```

#### Data Dictionary (Selected Variables)

The variables are grouped based on the dataset used for each model.

**1. Binary Classification Dataset (Marijuana Use)**

MRJFLAG – Marijuana use (0 = No, 1 = Yes)  
YOSELL2 – Involvement in selling-related activities  
YOFIGHT2 – Involvement in physical fights  
IRSEX – Sex (1 = Male, 2 = Female)  
INCOME – Household income level  
EDUSCHLGO – School enrollment status  
AVGGRADE – Average grades in school  
FRDMEVR2 – Friends have ever used marijuana  
PARCHKHW – Parents check homework  
RLGATTD – Religious service attendance  

**2. Multi-class Classification Dataset (Marijuana Levels)**

STNDSMJ – Level of marijuana use (categorical levels such as low, medium, high)  
IRSEX – Sex (1 = Male, 2 = Female)  
INCOME – Household income level  
EDUSCHLGO – School enrollment status  
AVGGRADE – Average grades in school  
YOFIGHT2 – Involvement in physical fights  
FRDMEVR2 – Friends have ever used marijuana  
PARCHKHW – Parents check homework  
RLGATTD – Religious service attendance  

**3. Regression Dataset (Alcohol Use)**

ALCYDAYS – Number of days alcohol was used  
IRSEX – Sex (1 = Male, 2 = Female)  
INCOME – Household income level  
EDUSCHLGO – School enrollment status  
AVGGRADE – Average grades in school  
YOFIGHT2 – Involvement in physical fights  
FRDMEVR2 – Friends have ever used marijuana  
PARCHKHW – Parents check homework  
RLGATTD – Religious service attendance  

---

### Renaming the variables

```{r}
youth_data <- youth_data %>%
  rename(
    Marijuana_use = MRJFLAG,
    Selling_behavior = YOSELL2,
    Physical_fight = YOFIGHT2,
    Sex = IRSEX,
    Race = NEWRACE2,
    Household_income = INCOME,
    School_enrollment = EDUSCHLGO,
    Average_grades = AVGGRADE,
    Friends_used_marijuana = FRDMEVR2,
    Parent_checks_homework = PARCHKHW,
    Religious_attendance = RLGATTD,
    Marijuana_use_level = STNDSMJ,
    Alcohol_use_days = ALCYDAYS,
    Parent_helps_homework = PARHLPHW,
  )
```

---

### Data Cleaning

Select Variables for Analysis 

```{r}
youth_model_data <- youth_data[, c(
  "Marijuana_use",
  "Marijuana_use_level",
  "Alcohol_use_days",
  "Sex",
  "Race",
  "Household_income",
  "School_enrollment",
  "Average_grades",
  "Parent_checks_homework",
  "Parent_helps_homework",
  "Physical_fight",
  "Selling_behavior",
  "Friends_used_marijuana",
  "Religious_attendance"
)]

# check structure
str(youth_model_data)
```

```{r}
# Handle Missing Values remove incomplete observations for modeling

youth_model_data <- na.omit(youth_model_data)

# check the size of the cleaned dataset
dim(youth_model_data)

youth_model_data[youth_model_data %in% c(91, 93, 94, 97, 98)] <- NA

# check how many missing values were created
colSums(is.na(youth_model_data))
```

The original dataset contained 10,561 observations and 79 variables. For this analysis, 14 relevant variables were selected based on the modeling goals. After removing incomplete observations, the final dataset contained 9,199 complete cases. This means 1,362 observations were excluded during data cleaning.

```{r}
# Check Class Distribution check how many users vs non-users

table(youth_model_data$Marijuana_use)

# check proportions (class imbalance)
prop.table(table(youth_model_data$Marijuana_use))
```

**The dataset is imbalanced (more non-users than users) this may affect model performance and interpretation**

---

### 1. Binary Classification (Marijuana_use)
The predictors include demographic, behavioral, and social variables such as sex, race, household income, school enrollment, academic performance, parental involvement, physical fighting, selling behavior, peer marijuana use, and religious attendance. The variables Marijuana_use_level and Alcohol_use_days were excluded to avoid using related outcome information.The goal of this model is to predict whether an individual has used marijuana (yes/no).

#### Model Setup

```{r}
# Train/Test Split
set.seed(123)

train_index <- createDataPartition(
  youth_model_data$Marijuana_use,
  p = 0.7,
  list = FALSE
)

train_data <- youth_model_data[train_index, ]
test_data  <- youth_model_data[-train_index, ]

# check sizes
dim(train_data)
dim(test_data)
```
After splitting, the training dataset contains 6,440 observations and the testing dataset contains 2,759 observations. These datasets will be used to train the model and then evaluate its performance on unseen data.

```{r}
# Cross-Validation Setup use 10-fold cross-validation for tuning

ctrl <- trainControl(
  method = "cv",
  number = 10
)

# Use 10-fold cross-validation to get a more reliable model evaluation
```

#### Hyperparameter Tuning

```{r}
# Decision Tree Tuning Grid test different cp values to find the best tree complexity

tree_grid <- expand.grid(
  cp = seq(0.001, 0.02, by = 0.002)
)

tree_grid
```

In this step, I created a tuning grid to test different `cp` values for the decision tree. The `cp` parameter controls the complexity of the tree, where smaller values allow more splits. I tested values from 0.001 to 0.02 to find the best balance between overfitting and underfitting.

```{r}
# Tune Decision Tree Model use cross-validation to find the best cp value

set.seed(123)

tree_cv <- train(
  Marijuana_use ~ .,
  data = train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))],
  method = "rpart",
  trControl = ctrl,
  tuneGrid = tree_grid
)

tree_cv
```
The model was evaluated using 10-fold cross-validation, and accuracy was used to select the best parameter. The optimal value was cp = 0.017, which gave the highest accuracy among all tested values.

```{r}
# Get the Best cp Value extract the optimal cp value from cross-validation results

best_cp <- tree_cv$bestTune$cp
best_cp
```

#### Final Model Fitting

```{r}
# Fit Final Decision Tree Model fit the final tree using the best cp value

tree_model <- rpart(
  Marijuana_use ~ .,
  data = train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))],
  method = "class",
  cp = best_cp
)
```

#### Model Evaluation

```{r}
# Evaluate Decision Tree Model on test data

tree_pred <- predict(tree_model, test_data, type = "class")

# Calculate accuracy
tree_accuracy <- mean(tree_pred == test_data$Marijuana_use)
tree_accuracy

# Confusion matrix
confusionMatrix(tree_pred, test_data$Marijuana_use)
```
The decision tree was evaluated using the test data. The model achieved an accuracy of about 84.8%, indicating that most observations were classified correctly.

However, the confusion matrix shows that the model predicts the majority class (non-users) very well, but struggles to correctly identify users. This is reflected in the low specificity and low kappa value, suggesting that class imbalance impacts the model’s performance.

#### Model Interpretation

```{r}
# Plot Final Decision Tree display the final decision tree

rpart.plot(tree_model)
```

The decision tree shows that the first and most important split is based on whether friends use marijuana. If friends do not use marijuana, most individuals are classified as non-users.

For those whose friends use marijuana, the model further splits based on selling behavior. Individuals involved in selling-related activities are more likely to be classified as users.

This suggests that peer influence and risky behavior play a key role in predicting marijuana use.
---

### 2. Random Forest Model:Marijuana Use

#### Hyperparameter Tuning

```{r}
# Random Forest Tuning Grid test different mtry values (number of variables at each split)

rf_grid <- expand.grid(
  mtry = 2:6
)

rf_grid
```
In this step, a tuning grid was created to test different mtry values for the random forest model. The mtry parameter controls how many variables are considered at each split. Values from 2 to 6 were tested to find the optimal setting that improves model performance.

```{r}
# Tune Random Forest Model use a smaller grid and fewer trees to reduce run time

set.seed(123)

rf_grid <- expand.grid(
  mtry = 2:4
)

rf_cv <- train(
  Marijuana_use ~ .,
  data = train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))],
  method = "rf",
  trControl = ctrl,
  tuneGrid = rf_grid,
  ntree = 100
)

rf_cv
```
Cross-validation was used to tune the random forest model by testing different mtry values. The model was evaluated using 10-fold cross-validation, and accuracy was used to select the best parameter. The optimal value was mtry = 4, which achieved the highest accuracy among the tested values.

```{r}
# Get the Best mtry Value extract the optimal mtry value from cross-validation results

best_mtry <- rf_cv$bestTune$mtry
best_mtry
```

#### Final Model Fitting

```{r}
# Fit Final Random Forest Model fit the final random forest using the best mtry

rf_model <- randomForest(
  Marijuana_use ~ .,
  data = train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))],
  mtry = best_mtry,
  ntree = 100
)
```

#### Model Evaluation

```{r}
# Evaluate Random Forest Model make predictions on test data

rf_pred <- predict(rf_model, test_data)

# calculate accuracy
rf_accuracy <- mean(rf_pred == test_data$Marijuana_use)
rf_accuracy

# confusion matrix
confusionMatrix(rf_pred, test_data$Marijuana_use)

varImpPlot(rf_model)
```

#### Model Interpretation

the random forest model was evaluated using the test data. The model achieved an accuracy of about 84.8%, similar to the decision tree.

The confusion matrix shows improved performance in identifying users compared to the decision tree, as reflected in higher specificity and kappa. However, the model still performs better at predicting the majority class, indicating that class imbalance continues to affect the results.

---

### 3. Boosting Model:Marijuana Use

#### Hyperparameter Tuning

```{r}
# Boosting Tuning Grid define parameter grid for boosting (gbm)

boost_grid <- expand.grid(
  n.trees = c(50, 100),
  interaction.depth = c(1, 2),
  shrinkage = c(0.01, 0.1),
  n.minobsinnode = 10
)

boost_grid
```
Tuning grid was created for the boosting model by testing different combinations of parameters. These include the number of trees, tree depth, and learning rate. This allows the model to explore different levels of complexity and find the best combination for performance.

```{r}
# Tune Boosting Model use cross-validation to find the best boosting parameters

set.seed(123)

boost_cv <- train(
  Marijuana_use ~ .,
  data = train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))],
  method = "gbm",
  trControl = ctrl,
  tuneGrid = boost_grid,
  verbose = FALSE
)

boost_cv
```
Cross-validation was used to tune the boosting model by testing different parameter combinations. The model was evaluated using 10-fold cross-validation, and accuracy was used to select the best settings. The optimal parameters included a higher learning rate and deeper trees, which provided the best performance among the tested options.

```{r}
# Get the Best Boosting Parameters extract the optimal tuning values from cross-validation results

best_n_trees <- boost_cv$bestTune$n.trees
best_depth <- boost_cv$bestTune$interaction.depth
best_shrinkage <- boost_cv$bestTune$shrinkage
best_n_minobs <- boost_cv$bestTune$n.minobsinnode

boost_cv$bestTune
```

The optimal boosting parameters were extracted from the cross-validation results. The best combination included 50 trees, a tree depth of 2, and a learning rate of 0.1. These values were selected because they achieved the highest accuracy among all tested combinations.

#### Final Model Fitting

```{r}
# Convert Response Variable for Boosting gbm with Bernoulli requires the outcome to be numeric (0 and 1)

train_boost <- train_data[, -which(names(train_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))]
test_boost  <- test_data[, -which(names(test_data) %in% c("Marijuana_use_level", "Alcohol_use_days"))]

train_boost$Marijuana_use <- as.numeric(as.character(train_boost$Marijuana_use))
test_boost$Marijuana_use  <- as.numeric(as.character(test_boost$Marijuana_use))
```

```{r}
# Fit Final Boosting Model fit the final boosting model using the best tuning parameters

boost_model <- gbm(
  Marijuana_use ~ .,
  data = train_boost,
  distribution = "bernoulli",
  n.trees = best_n_trees,
  interaction.depth = best_depth,
  shrinkage = best_shrinkage,
  n.minobsinnode = best_n_minobs,
  verbose = FALSE
)
```

#### Model Evaluation

```{r}
# Evaluate Boosting Model predict probabilities and convert them to class labels

boost_prob <- predict(boost_model, test_boost, n.trees = best_n_trees, type = "response")
boost_pred <- ifelse(boost_prob > 0.5, 1, 0)

# calculate accuracy
boost_accuracy <- mean(boost_pred == test_boost$Marijuana_use)
boost_accuracy

# confusion matrix
confusionMatrix(as.factor(boost_pred), as.factor(test_boost$Marijuana_use))
```

#### Model Interpretation

The boosting model was evaluated using the test data. The model achieved an accuracy of about 84.9%, which is slightly higher than the decision tree and random forest.

The confusion matrix shows some improvement in identifying users, but the model still performs better on the majority class. This suggests that class imbalance continues to influence the model’s performance.

---

### 4. Multi-class Classification (Marijuana_use_level)
The same set of demographic, behavioral, and social predictors was used. The variables Marijuana_use and Alcohol_use_days were excluded, since Marijuana_use_level is the response variable and Marijuana_use is closely related to it. The goal is to classify individuals into different levels of marijuana use.

#### Model Setup

```{r}
# Multi-Class Decision Tree Setup create datasets for predicting marijuana use level

train_multi <- train_data[, -which(names(train_data) %in% c("Marijuana_use", "Alcohol_use_days"))]
test_multi  <- test_data[, -which(names(test_data) %in% c("Marijuana_use", "Alcohol_use_days"))]

# make sure the outcome is a factor
train_multi$Marijuana_use_level <- as.factor(train_multi$Marijuana_use_level)
test_multi$Marijuana_use_level  <- as.factor(test_multi$Marijuana_use_level)
```

#### Model Setup

```{r}
# Multi-Class Decision Tree Tuning Grid test different cp values for the multi-class tree

multi_grid <- expand.grid(
  cp = seq(0.001, 0.02, by = 0.002)
)
```

```{r}
# Tune Multi-Class Decision Treeuse cross-validation to find the best cp value

set.seed(123)

multi_cv <- train(
  Marijuana_use_level ~ .,
  data = train_multi,
  method = "rpart",
  trControl = ctrl,
  tuneGrid = multi_grid
)

multi_cv
```

Cross-validation was used to tune the multi-class decision tree by testing different cp values. The model was evaluated using 10-fold cross-validation, and accuracy was used to select the best parameter. The optimal value was cp = 0.001, which achieved the highest accuracy among the tested values.

```{r}
# Get the Best cp Value extract the optimal cp value

best_multi_cp <- multi_cv$bestTune$cp
best_multi_cp
```

#### Final Model Fitting

```{r}
# Fit Final Multi-Class Decision Tree fit the final multi-class tree using the best cp value

multi_model <- rpart(
  Marijuana_use_level ~ .,
  data = train_multi,
  method = "class",
  cp = best_multi_cp
)
```

#### Model Evaluation

```{r}
# Evaluate Multi-Class Decision Tree make predictions on the test data

multi_pred <- predict(multi_model, test_multi, type = "class")

# calculate accuracy
multi_accuracy <- mean(multi_pred == test_multi$Marijuana_use_level)
multi_accuracy

# confusion matrix
confusionMatrix(multi_pred, test_multi$Marijuana_use_level)
```

#### Model Interpretation

The multi-class decision tree was evaluated using the test data. The model achieved an accuracy of about 76.2%, which is lower than the binary classification models.

The confusion matrix shows that the model performs better at predicting one class compared to the other, indicating some imbalance in classification performance. This suggests that predicting multiple levels of marijuana use is more complex than binary classification.

```{r}
# Plot Final Multi-Class Decision Tree display the final multi-class decision tree

rpart.plot(multi_model)
```

The multi-class decision tree shows multiple splits based on behavioral and social factors. Compared to the binary tree, this model is more complex, with deeper branches used to separate different levels of marijuana use.

The tree indicates that variables such as peer influence and behavior continue to play an important role, but the patterns are less clear. This reflects the difficulty of distinguishing between multiple usage levels rather than simply identifying use or non-use.

---

### 5.Regression (Alcohol_use_days)
The predictors again include the same demographic, behavioral, and social variables. The variables Marijuana_use and Marijuana_use_level were excluded to ensure that alcohol use is predicted independently without using related marijuana outcomes. The goal is to predict the number of days of alcohol use as a continuous outcome.

#### Model Setup

```{r}
# Regression Tree Setup create datasets for predicting alcohol use days

train_reg <- train_data[, -which(names(train_data) %in% c("Marijuana_use", "Marijuana_use_level"))]
test_reg  <- test_data[, -which(names(test_data) %in% c("Marijuana_use", "Marijuana_use_level"))]
```

#### Hyperparameter Tuning

```{r}
# Regression Tree Tuning Grid test different cp values for the regression tree

reg_grid <- expand.grid(
  cp = seq(0.001, 0.02, by = 0.002)
)
```

```{r}
# Tune Regression Tree use cross-validation to find the best cp value

set.seed(123)

reg_cv <- train(
  Alcohol_use_days ~ .,
  data = train_reg,
  method = "rpart",
  trControl = ctrl,
  tuneGrid = reg_grid
)

reg_cv
```

Cross-validation was used to tune the regression tree by testing different cp values. The model was evaluated using RMSE, and the best value was selected based on the lowest error. The optimal parameter was cp = 0.003, which produced the smallest RMSE among the tested values.

```{r}
# Get the Best cp Value xtract the optimal cp value for the regression tree

best_reg_cp <- reg_cv$bestTune$cp
best_reg_cp
```

#### Final Model Fitting

```{r}
# Fit Final Regression Tree fit the final regression tree using the best cp value

reg_model <- rpart(
  Alcohol_use_days ~ .,
  data = train_reg,
  method = "anova",
  cp = best_reg_cp
)
```

#### Model Evaluation

```{r}
# Evaluate Regression Tree make predictions on the test data

reg_pred <- predict(reg_model, test_reg)

# calculate mse
reg_mse <- mean((reg_pred - test_reg$Alcohol_use_days)^2)
reg_mse

# calculate rmse
reg_rmse <- sqrt(reg_mse)
reg_rmse
```
The regression tree was evaluated using the test data. The model produced an MSE of about 2.83 and an RMSE of about 1.68. These values represent the average error between the predicted and actual number of alcohol use days, showing that the model captures general trends but still has some prediction error.

#### Model Interpretation

```{r}
# Plot Final Regression Tree display the final regression tree

rpart.plot(reg_model)
```

The regression tree shows that the main split is based on whether friends use marijuana. This indicates that peer influence is an important factor in predicting alcohol use days.

For individuals whose friends use marijuana, the predicted number of alcohol use days is lower, with further differences based on household income and selling behavior. For those whose friends do not use marijuana, higher alcohol use is observed, with additional splits based on selling behavior and parental involvement.

This suggests that both social environment and family factors influence alcohol use patterns.

---

#### Results

For binary classification, the decision tree achieved a test accuracy of 0.8481, the random forest also achieved 0.8481, and boosting performed slightly better with an accuracy of 0.8492. However, these results should be interpreted carefully because the marijuana-use outcome is imbalanced, with about 84% non-users in the cleaned dataset. This means that accuracy by itself can overstate model quality if a model mainly predicts the majority class. Therefore, the binary results suggest that the models perform reasonably well overall, but the improvement over a simple majority-class prediction is limited. In the multi-class task, the decision tree achieved an accuracy of 0.7619, which is lower than the binary models because predicting more than two categories is a more difficult problem. In the regression task, the model produced an MSE of 2.8269 and an RMSE of 1.6813. The RMSE is easier to interpret in context: on average, the predicted number of alcohol-use days is off by about 1.68 days. Overall, boosting had the best binary accuracy, but the difference from the other models was very small, so the practical advantage was limited.

### Model Comparison

```{r}
# # Create Model Performance Table combine model results into one summary table

performance_table <- data.frame(
  Model = c(
    "Decision Tree",
    "Random Forest",
    "Boosting",
    "Multi-Class Tree",
    "Regression Tree"
  ),
  Metric = c(
    "Accuracy",
    "Accuracy",
    "Accuracy",
    "Accuracy",
    "MSE"
  ),
  Value = c(
    tree_accuracy,
    rf_accuracy,
    boost_accuracy,
    multi_accuracy,
    reg_mse
  )
)

performance_table

# Format Model Performance Table round values for easier presentation

performance_table$Value <- round(performance_table$Value, 3)

performance_table
```


When comparing the models, boosting gave the highest binary classification accuracy, but only by a very small margin over the decision tree and random forest. This suggests that the more complex ensemble models did not dramatically outperform the simpler tree for this dataset. The decision tree remains the easiest to interpret because its splits can be directly explained, while the random forest and boosting models provide better flexibility at the cost of interpretability. The multi-class tree had lower accuracy than the binary models, which is expected because distinguishing between multiple marijuana-use levels is more complex than predicting use versus non-use. For regression, the tree provides a reasonable summary of alcohol-use patterns, but prediction error still exists, with an average error of about 1.68 days. Taken together, these results show that tree-based methods can be useful for different outcome types, but model performance should always be interpreted in light of class imbalance, error size, and the practical meaning of the predictions.

---

### Ethical Considerations

When interpreting the results, it is important not to make assumptions about individuals or groups. The model shows patterns in the data, but this does not mean it proves cause and effect.

For example, demographic or behavioral variables may be linked to substance use, but they should not be used to label or stereotype people. Misunderstanding these results can lead to unfair conclusions. Furthermore, risk of misusing model predictions in decision-making. For instance, if these results were used by schools or organizations, individuals labeled as “high risk” could be unfairly targeted or monitored. This may lead to unequal treatment.

Therefore, the results should be used carefully, focusing on general trends rather than judging specific individuals or groups.